On the computational complexity of detecting possibilistic locality

### 2017-06-15 paper.tex
Introduction: Rewrote paragraph beginning "In this paper" to give a more concrete background to what a Bell experiment is.

\section{Introduction}
Since the inception of quantum mechanics, the apparent nonlocality of the theory has been a thorny issue. Einstein, together with Podolsky and Rosen \cite{EPR}, argued that this property, in apparent contravention of the information-propagation bound of the speed of light from special relativity, was evidence of the incompleteness of the quantum-mechanical description of nature. The related concept of nonlocal steering was also the subject of Einstein's scrutiny in a letter to Edwin Schr\"{o}dinger \cite{Eins1935}, in which it is noted that a local description of steering necessitates nature to have what we would now call the $\psi$-epistemic property: that the quantum state is not instantiated as a part of nature's ontology.

In this paper, we will consider ``Bell experiments'' of the classic form: we have two separate quantum systems, in a possibly entangled state, and a set of observables that we can measure on each. Formally, this is a prepare-and-measure scenario for which the commutativity graph of the available observables is bipartite; these notions are equivalent since we will consider systems of finite quantum dimension. Such an experiment will have probabilities for its outcomes predicted by quantum theory; we are then interested in whether or not these predictions could be realised via a local (that is, factorisable) hidden variable model. That this is not always possible was first noted by Bell \cite{Bell1966}.

In 1991, Avis \emph{et al} \cite{Avis2004} demonstrated that determining whether or not a given table of these operational predictions for a Bell experiment had a local hidden variable explanation was \textbf{NP}-hard, building on work by Pitowsky \cite{Pito1992}. In this paper, we consider a more extreme, possibilistic, manifestation of nonlocality; one in which it is not only true that we cannot account for the specific experimental probabilities with a local hidden variable model, but that we cannot even account for the possibilities. In doing so, we will identify the minimal operational requirements for this problem to remain \textbf{NP}-Hard, filling in the final gap in a categorisation started by Mansfield and Fritz \cite{Mans2011,Mans2016} and continued by Abramsky, Gottlob and Kolaitis \cite{Abra2013}.

@@

\item For each intersection of a variable row and clause column, such that the variable is represented negatively in the clause, we have the measurement possibilities given by:
\begin{equation}
\begin{array}{  | c c  c|}
\hline
 1 & 1 & 1 \\
 0 & 1& 1 \\ \hline
 \end{array}\, ,
 \begin{array}{  | c c  c|}
\hline
 1 &1 & 1 \\
 1 & 0& 1 \\ \hline
 \end{array} \, ,
 \mbox{or} \,
 \begin{array}{  | c c  c|}
\hline
 1 & 1 & 1\\
 1 & 1& 0 \\ \hline
 \end{array}\, ,
 \end{equation}
depending on whether the variable is the first, second, or third variable in the clause with respect to our variable ordering.
\end{itemize}
We can see that choices of subrow in a measurement row is equivalent to an assignment of that variable, and that the clause structure effectively bans the assignment disallowed by that specific clause, for example the clause $(x_1 \vee x_2 \vee x_3)$ corresponds to the possibility table
\begin{equation*}
\begin{array}{c  | c c  c|}
\hline
 x_1=0&0 & 1 & 1 \\
 x_1=1&1 & 1& 1 \\ \hline
 x_2=0&1 &0 & 1 \\
x_2=1& 1 & 1& 1 \\ \hline
x_3=0& 1 & 1 & 0\\
x_3=1& 1 & 1& 1 \\ \hline
 \end{array}
 \end{equation*}
We see we cannot simultaneously assign a 1 to each of the top subrows as part of a deterministic grid. We, then, need only to answer the question of the necessary and sufficient nature of \textsc{2-Robust}ness; we note that choosing a specific 1 to complete to a grid represents therefore an assignment of one variable, and the structure of the table then might result in the forced selection of another variable's assignment, if there is a 0 in the same subcolumn as the selected 1. Therefore by the same logic as before, the \textsc{(2,3)-PossLoc} decision problem is \textbf{NP}-complete.
\end{proof}
Therefore, since \textsc{2-Robust 3-SAT} is \textbf{NP}-Hard; any instance of \textsc{2-Robust 3-SAT} can be reduced to an instance of \textsc{(2,3)-PossLoc} in polynomial time; and \textsc{(2,3)-PossLoc} is in \textbf{NP}, \textsc{(2,3)-PossLoc} is \textbf{NP}-complete.
\end{proof}

>>>>

\section{Introduction}
Since the inception of quantum mechanics, the apparent nonlocality of the theory has been a thorny issue. Einstein, together with Podolsky and Rosen \cite{EPR}, argued that this property, in apparent contravention of the information-propagation bound of the speed of light from special relativity, was evidence of the incompleteness of the quantum-mechanical description of nature. The related concept of nonlocal steering was also the subject of Einstein's scrutiny in a letter to Edwin Schr\"{o}dinger \cite{Eins1935-6-19}, in which it is noted that a local description of steering necessitates nature to have what we would now call the $\psi$-epistemic property: that the quantum state is not instantiated as a part of nature's ontology.

In this paper, we will consider ``Bell experiments'' of the classic form. Two experimenters have laboratories in separated locations, performing measurements on two quantum systems; each experimenter has a variety of measurements at their disposal, and once the measurements are done, they compare results. Quantum theory provides a framework in which the probability of the different outcomes, given the measurement settings chosen by each experimenter, can be calculated. In 1966, Bell showed that in some cases, the statistics predicted by quantum mechanics was incompatible with a notion of local realism; that each experimenter was performing measurements on a system with a definite state, independent of the measurement choices made at the other location \cite{Bell1966} When such an explanation in terms of local hidden variables is not possible, we would say that the distribution is \emph{nonlocal}, or demonstrates \emph{Bell nonlocality}. This motivates a natural decision problem: we are given a table of empirical data from such a Bell experiment, and asked to decide whether or not it constitutes a demonstration of nonlocality.

In 1991, Avis \emph{et al} \cite{Avis2004} demonstrated that determining whether or not a given table of these operational predictions for a Bell experiment had a local hidden variable explanation was \textbf{NP}-hard, building on work by Pitowsky \cite{Pito1991}. In this paper, we consider a more extreme, possibilistic, manifestation of nonlocality; one in which it is not only true that we cannot account for the specific experimental probabilities with a local hidden variable model, but that we cannot even account for the possibilities. In doing so, we will identify the minimal operational requirements for this problem to remain \textbf{NP}-hard, filling in the final gap in a categorisation started by Mansfield and Fritz \cite{Mans2011,Mans2016} and continued by Abramsky, Gottlob and Kolaitis \cite{Abra2013}.

@@

\item For each intersection of a variable row and clause column, such that the variable is represented negatively in the clause, we have the measurement possibilities given by:
\begin{equation}
\begin{array}{  | c c  c|}
\hline
 1 & 1 & 1 \\
 0 & 1& 1 \\ \hline
 \end{array}\, ,
 \begin{array}{  | c c  c|}
\hline
 1 &1 & 1 \\
 1 & 0& 1 \\ \hline
 \end{array} \, ,
 \mbox{or} \,
 \begin{array}{  | c c  c|}
\hline
 1 & 1 & 1\\
 1 & 1& 0 \\ \hline
 \end{array}\, ,
 \end{equation}
depending on whether the variable is the first, second, or third variable in the clause with respect to our variable ordering.
\end{itemize}
We can see that choices of subrow in a measurement row is equivalent to an assignment of that variable, and that the clause structure effectively bans the assignment disallowed by that specific clause, for example the clause $(x_1 \vee x_2 \vee x_3)$ corresponds to the possibility table
\begin{equation*}
\begin{array}{c  | c c  c|}
\hline
 x_1=0&0 & 1 & 1 \\
 x_1=1&1 & 1& 1 \\ \hline
 x_2=0&1 &0 & 1 \\
x_2=1& 1 & 1& 1 \\ \hline
x_3=0& 1 & 1 & 0\\
x_3=1& 1 & 1& 1 \\ \hline
 \end{array}
 \end{equation*}
We see we cannot simultaneously assign a 1 to each of the top subrows as part of a deterministic grid. We, then, need only to answer the question of the necessary and sufficient nature of \textsc{2-Robust}ness; we note that choosing a specific 1 to complete to a grid represents therefore an assignment of one variable, and the structure of the table then might result in the forced selection of another variable's assignment, if there is a 0 in the same subcolumn as the selected 1. Therefore by the same logic as before, the \textsc{(2,3)-PossLoc} decision problem is \textbf{NP}-complete.
\end{proof}
Therefore, since \textsc{2-Robust 3-SAT} is \textbf{NP}-hard; any instance of \textsc{2-Robust 3-SAT} can be reduced to an instance of \textsc{(2,3)-PossLoc} in polynomial time; and \textsc{(2,3)-PossLoc} is in \textbf{NP}, \textsc{(2,3)-PossLoc} is \textbf{NP}-complete.
\end{proof}
